Postural developments mediate children's visual access to social information

### 2018-01-28 writeup-cogsci/cogsci2018_faceOnly.tex
more edits

\begin{abstract}
The ability to process social cues--including eye gaze--is a critical
component of children's early language and cognitive development.
However, as children reach their first birthday, they begin to locomote
themselves, walking and exploring their visual environment in an
entirely new way. How do these postural and locomotive changes affect
children's access to the social information relevant for word-learning?
Here, we explore this question by using head-mounted cameras to record
infants' (8-16 months of age) egocentric visual perspective and use
computer vision algorithms to detect the proportion of faces in infants'
environments. We find that infants' posture and orientation to their
caregiver largely mediate infants' access to faces, suggesting that
these postural and locomotive developments facilitate infants' emerging
linguistic and social capacities. Broadly, we suggest that the combined
use of head-mounted cameras and the application of new computer vision
techniques is a promising avenue for understanding the statistics of
infants' visual and linguistic experience.

@@

One possibility is that these motor improvements have strong
developmental cascades, impacting children's emerging social, cognitive,
and linguistic abilities (Iverson, 2010). Indeed, these postural changes
also impact how children interact with their mothers; walking
(vs.~crawling) infants make different kinds of object-related bids for
attention from their mothers and tend to hear more action directed
statements (e.g., ``open it'') (Karasik, Tamis-LeMonda, \& Adolph,
2014). Further, in an observational study, Walle \& Campos (2014) found
that children who were able to walk had both higher receptive and
productive vocabularies. On their account, children's ability to stand
and independently locomote may fundamentally change their ability to
access social information (e.g., faces, gaze) relative to children who
are still crawling and sitting. In other words, the ability to walk and
move around independently and subsequently access more detailed social
information may accelerate infants' ability to learn the people in their
environments.

@@

Here, we directly ask whether postural and locomotive developments
change the availability of social information--the presence of faces and
hands. To do so, we recorded the visual experience of a group of infants
in three age ranges (8,12, and 16 months) using head-mounted cameras
during a brief laboratory free-play session; children's posture and
orientation relative to their caregiver were also recorded from a
third-person perspective and hand-annotated. We then capitalize on
recent improvements in face detection algorithms (Cao, Simon, Wei, \&
Sheikh, 2017; K. Zhang, Zhang, Li, \& Qiao, 2016) to analyze the
frequencies of faces and wrists/hands in the child's visual environment,
both overall and relative to naming events by their caregivers. We
hypothesized that there would be differential access to social
information based on children's postural developments: that as infant's
learned how to crawl, they would see relatively fewer faces as they
would be looking primarily at the ground, and that as infant's learned
how to walk and locomote independently, that they will experience a much
greater portion of social information in their environment.

@@

\caption[Field of view for two different headcam configurations]{Field of view for two different headcam configurations. We used the second device in our experiments.}\label{fig:headcam}
\end{figure}
\end{CodeChunk}

@@

We created a set of custom annotations that described the child's
physical posture (e.g.~standing) and the orientation of the caregiver
relative to the child (e.g.~far away). The child's posture was
categorized as being held/carried, prone (crawling or lying), sitting,
or standing. The caregiver's orientation was characterized as being
close to the child, far from the child, and behind the child (whether
close or far). For the first two annotations (close/far from the child),
the caregiver could either be to the the front or to the side of the
child. All annotations were made by a trained coder using the
OpenSHAPA/Datavyu software (K. Adolph, Gilmore, Freeman, Sanderson, \&
Millman, 2012), and times when the child was out of view of the tripod
camera were marked as uncodable and were excluded from these
annotations.

\subsection{Face Detection}\label{face-detection}

We used three face detection systems to explore infants' changing access
to social information and to avoid the cost of hand-annotating every
frame. We first measured the performance of the most commonly-used and
widely available face detection algorithm, Viola-Jones. We used this as
a benchmark for performance, and while it can achieve impressive
accuracy in some situations, it is notoriously bad at dealing with
occluded faces (Scheirer, Anthony, Nakayama, \& Cox, 2014). We next
capitalized on recent improvements in computer vision, testing the
performance of two face detectors that both made use of Convolutional
Neural Networks (CNNs) to extract face information. The first algorithm
was specifically optimized for face detection, and the second algorithm
was optimized to extract information about the position of agent's
bodyparts.

The second of these algorithms (OpenPose) also captures information
about the ``skeleton'' of people in a given frame, outputting
information about the entire set of body parts that were detected. We
analyze wrist detections as a good proxy for the presence of a hand. We
do so first because we did not capture the entire visual field
experienced by the infants/children. Thus, even though the video may
only show the presence of a wrist, the caregiver's hand may be within
the child's field of view. Second, we do so because hands are often
occluded by objects when caregivers are interacting with children, yet
still visually accessible by the child and part of their joint
interaction. For example, if a caregiver was holding a toy and
presenting it to the infant, their wrists may be visible but their hand
may not necessarily be (as it is occluded by the toy).

\subsubsection{Face Detection
Algorithms}\label{face-detection-algorithms}

@@

The third algorithm was a CNN-based pose detector that provided the
locations of 18 body parts (ears, nose, wrists, etc.) called OpenPose
(Cao et al., 2017; Simon, Joo, Matthews, \& Sheikh, 2017; Wei,
Ramakrishna, Kanade, \& Sheikh, 2016) avaliable at
\url{https://github.com/CMU-Perceptual-Computing-Lab/openpose}. The
system uses a CNN for initial anatomical detection and subsequently
applies part affinity fields (PAFs) for part association, producing a
series of body part candidates. The candidates are then matched to a
single individual and finally assembled into a pose. For the purposes of
our investigation we only made use of the body parts relevant to the
face (ears, eyes, nose). We operationalized face detections as any
frames containing a nose, as any half of a face necessarily contains a
nose. We operationalized wrist detections as any frames containing
either the left or the right wrist.

@@

To evaluate face detector performance, we needed to construct a ``gold
set'' of labeled frames that did not carry any biases that might be due
to the relatively rare appearance of faces in the dataset. We thus
hand-labeled two samples: a sample containing a high density of faces
(half reported by MTCNN, half by OpenPose) and a random sample from the
remaining frames. Each sample was comprised of an equal number of frames
taken from each participant (i.e., video). Faces were classifed as
present in a frame if at least half of the face was showing. Precision
(hits / hits + false alarms), recall (hits / hits + misses), and F-score
(harmonic mean of precision and recall) were calculated for all
detectors and are reported in Table 2. For wrist detections, the ``gold
set'' was constructed in the same way, except that frames with a high
density of wrists were taken by wrist detections made by OpenPose.

For face detection, MTCNN outperformed OpenPose when looking at the
composite F-score (0.89 vs.~0.83); thus the face presence analyses to
follow all result from MTCNN detections. We found that although MTCNN
and OpenPose perform comparably with the random sample, MTCNN had the
edge with the high density sample (specifically looking at precision),
suggesting that OpenPose generated more false positives than MTCNN.
ViolaJones performed quite poorly, missing many of the faces in the
randomly sampled frames.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rllrrr}
  \hline
   Algorithm & Sample\ Type & P & R & F \\ 
  \hline
  MTCNN-Faces & High density & 0.89 & 0.92 & \textbf{0.90} \\ 
  MTCNN-Faces & Random & 0.94 & 0.62 & 0.75 \\ 
  OpenPose-Faces & High density & 0.78 & 0.93 & 0.84 \\ 
  OpenPose-Faces & Random & 0.72 & 0.80 & \textbf{0.76} \\ 
  ViolaJones-Faces & High density & 0.96 & 0.44 & 0.60 \\ 
  ViolaJones-Faces & Random & 0.44 & 0.38 & 0.41 \\ 
  OpenPose-Wrists & High density & 0.64 & 1.00 & 0.78 \\ 
  OpenPose-Wrists & Random & 0.76 & 0.26 & 0.39 \\ 
   \hline
\end{tabular}
\caption{Detector performance on both high density samples (where proportion of faces detected was high) and random samples (where frames were randomly selected). P, R, and F denote precision, recall, and F-score, respectively.} 
\end{table}

@@

First, we report developmental shifts in infants' posture and their
orientation relative to their caregiver. Then, we explore how these
changes influence children's visual access to faces across this
developmental time range. Finally, we explore how these changes impact
the accessibility of faces during labeling events.

@@

\section{Changes in Access to Faces}\label{changes-in-access-to-faces}

@@

\caption[Proportion of faces detected by the MTCNN model (left) and wrists detected by the OpenPose model (right panel) as a function of child's age]{Proportion of faces detected by the MTCNN model (left) and wrists detected by the OpenPose model (right panel) as a function of child's age. Larger dots indicate children who had longer playsessions and thus for whom there was more data}\label{fig:detByAge}
\end{figure}
\end{CodeChunk}

We first examined the proportion of face detections across age; a full
summary can be seen in Figure \ref{fig:detByAge}. We observed a slight
U-shaped function both when analyzing the output of the MTCNN and
OpenPose detectors, such that 12-month-olds appeared to experience
slighly fewer faces faces than 8 or 16-month-olds.

@@

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
Intercept & -5.2469 & 0.0584 & -89.88 & 0.0000 \\ 
  Age & 0.0847 & 0.0041 & 20.89 & 0.0000 \\ 
  Prone & 0.2015 & 0.0564 & 3.57 & 0.0004 \\ 
  Sit & 1.4053 & 0.0541 & 25.98 & 0.0000 \\ 
  Stand & 1.4272 & 0.0542 & 26.33 & 0.0000 \\ 
  Close & 1.8239 & 0.0230 & 79.17 & 0.0000 \\ 
  Far & 2.5479 & 0.0239 & 106.42 & 0.0000 \\ 
   \hline
\end{tabular}
\caption{Model coefficients from a generalized linear model predicting the proportion of wrists seen by infants.} 
\end{table}

@@

To formalize these observations, we fit a generalized linear model to
the proportion of faces infants saw in each posture and orientation (as
detected by MTCNN), with participant's age, orientation, and posture as
independent variables. A summary of the coefficients of a model with
only main effects (and no interactions) model can be found in Table 3.
When we did include interaction terms between age, posture, and
orientation, age no longer remained a significant predictor (b = -15.84,
SE = 113.98, z = -0.14 , p =0.89). We found the same pattern of results
for the OpenPose face detections, omitted here for brevity. We also
found the same pattern with respect to infant's visual access to hands
(see Table 4); including interaction terms also reduced any main effect
of age on the proportion of hands detected (b = -15.7, SE = 113.98, z =
-0.14 , p =0.89). Thus, these results suggest that infants access to
social information is heavily infleunced by their postural and
locomotive development.

@@

Finally, we analyzed how face and wrists detections changed during
object labeling events as a function of infant's posture and
orientation. Specfically, we analyzed a four-second window around each
labeling event (e.g., ``Look at the {[}zem{]}!''); these labeling events
were hand-annotated and automatically synchronized with the
frame-by-frame face detections. We again found that infant's posture and
orientations impacted the degree to which they saw their caregiver's
face and wrist during a labeling event; infants who were sitting or
standing were more likely to have access to this social information.

>>>>

\begin{abstract}
The ability to process social cues--including eye gaze--is a critical
component of children's early language and cognitive development.
However, as children reach their first birthday, they begin to locomote
themselves, walking and exploring their visual environment in an
entirely new way. How do these postural and locomotive changes affect
children's access to the social information relevant for word-learning?
Here, we explore this question by using head-mounted cameras to record
infants' (8-16 months of age) egocentric visual perspective and use
computer vision algorithms to detect the proportion of faces and hands
in infants' environments. We find that infants' posture and orientation
to their caregiver largely mediate infants' access to socially
informative features, suggesting that postural and locomotive
development play a significant role in mediating children's emerging
linguistic and social capacities. Broadly, we suggest that the combined
use of head-mounted cameras and the application of new computer vision
techniques is a promising avenue for understanding the statistics of
infants' visual and linguistic experience.

@@

One possibility is that these motor improvements have strong
developmental cascades, impacting children's emerging social, cognitive,
and linguistic abilities (Iverson, 2010). Indeed, these postural changes
also impact how children interact with their mothers; walking
(vs.~crawling) infants make different kinds of object-related bids for
attention from their mothers and tend to hear more action directed
statements (e.g., ``open it'') (Karasik, Tamis-LeMonda, \& Adolph,
2014). Further, in an observational study, Walle \& Campos (2014) found
that children who were able to walk had both higher receptive and
productive vocabularies. On their account, children's ability to stand
and independently locomote may fundamentally change their ability to
access social information (e.g., faces, gaze) relative to children who
are still crawling and sitting. In other words, the ability to walk and
move around independently and subsequently access more detailed social
information may accelerate infants' ability to learn from the people in
their environments.

@@

Here, we directly ask whether postural and locomotive developments
change the availability of social information--the presence of faces and
hands. To do so, we recorded the visual experience of a group of infants
in three age ranges (8,12, and 16 months) using head-mounted cameras
during a brief laboratory free-play session; children's posture and
orientation relative to their caregiver were also recorded from a
third-person perspective and hand-annotated. We then capitalize on
recent improvements in face and pose detection algorithms (Cao, Simon,
Wei, \& Sheikh, 2017; K. Zhang, Zhang, Li, \& Qiao, 2016) to analyze the
frequencies of faces and hands (operationalized by the presence of
wrists) in the child's visual environment, both overall and relative to
naming events by their caregivers. We hypothesized that there would be
differential access to social information based on children's postural
developments: crawling infants would see fewer faces because they would
primarily be looking at the ground, and walking toddlers would have
access to a richer visual landscape that would contain a larger portion
of the social information in the environment.

@@

\caption[Field of view for two different headcam configurations (we used the one below in our current study)]{Field of view for two different headcam configurations (we used the one below in our current study).}\label{fig:headcam}
\end{figure}
\end{CodeChunk}

@@

We created a set of custom annotations that described the child's
physical posture (e.g.~standing) and the orientation of the caregiver
relative to the child (e.g.~far away). The child's posture was
categorized as being held/carried, prone (crawling or lying), sitting,
or standing. The caregiver's orientation was characterized as being
close to the child, far from the child, and behind the child
(independent of distance). For the first two annotations (close/far from
the child), the caregiver could either be to the the front or to the
side of the child. All annotations were made by a trained coder using
the OpenSHAPA/Datavyu software (K. Adolph, Gilmore, Freeman, Sanderson,
\& Millman, 2012), and times when the child was out of view of the
tripod camera were marked as uncodable and were excluded from these
annotations.

\subsection{Face and Wrist Detection}\label{face-and-wrist-detection}

We used three face detection systems to explore infants' changing access
to social information and to avoid the cost of hand-annotating every
frame. The first of these is the most commonly-used and widely available
face detection algorithm: Viola-Jones. We used this as a benchmark for
performance, and while it can achieve impressive accuracy in some
situations, it is notoriously bad at dealing with occluded faces
(Scheirer, Anthony, Nakayama, \& Cox, 2014). We next capitalized on
recent improvements in computer vision, testing the performance of two
face detectors that both made use of Convolutional Neural Networks
(CNNs) to extract face information. The first algorithm was specifically
optimized for face detection, and the second algorithm was optimized to
extract information about the position of 18 different body parts
belonging to an agent.

This second of the CNN-based algorithms - the pose detector - provided
us with the location of an agent's wrists, which we used as a good proxy
for the presence of a hand. We do so first because we did not capture
the entire visual field experienced by the infants/children. Thus, even
though the video may only show the presence of a wrist, the caregiver's
hand may be within the child's field of view. Second, we do so because
hands are often occluded by objects when caregivers are interacting with
children, yet still visually accessible by the child and part of their
joint interaction. For example, if a caregiver was holding a toy and
presenting it to the infant, their wrists may be visible but their hand
may not necessarily be (as it is occluded by the toy).

\subsubsection{Algorithms}\label{algorithms}

@@

The third algorithm was a CNN-based pose detector that provided the
locations of 18 body parts (ears, nose, wrists, etc.) called OpenPose
(Cao et al., 2017; Simon, Joo, Matthews, \& Sheikh, 2017; Wei,
Ramakrishna, Kanade, \& Sheikh, 2016) avaliable at
\url{https://github.com/CMU-Perceptual-Computing-Lab/openpose}. The
system uses a CNN for initial anatomical detection and subsequently
applies part affinity fields (PAFs) for part association, producing a
series of body part candidates. The candidates are then matched to a
single individual and finally assembled into a pose. For the purposes of
our investigation we only made use of the body parts relevant to the
face and hands (nose and wrists). We operationalized face detections as
any frames containing a nose, as any half of a face necessarily contains
a nose. We operationalized wrist detections as any frames containing
either the left or the right wrist.

@@

To evaluate face detector performance, we needed to construct an
unbiased ``gold set'' of labeled frames that made sure to account for
the relatively rare appearance of faces in the dataset. We thus
hand-labeled two samples: a sample containing a high density of faces
(half reported by MTCNN, half by OpenPose) and a random sample from the
remaining frames. Each sample was comprised of an equal number of frames
taken from each participant (i.e., video). Faces were classifed as
present in a frame if at least half of the face was showing. Precision
(hits / hits + false alarms), recall (hits / hits + misses), and F-score
(harmonic mean of precision and recall) were calculated for all
detectors and are reported in Table 2. For wrist detections, the ``gold
set'' was constructed in the same manner, with the exception that frames
with a high density of wrists were taken by wrist detections made by
OpenPose.

For face detection, MTCNN outperformed OpenPose when looking at the
composite F-score (0.89 vs.~0.83); thus the face presence analyses to
follow all result from MTCNN detections. We found that although MTCNN
and OpenPose perform comparably with the random sample, MTCNN performs
better on the high density sample evaluation (specifically looking at
precision), suggesting that OpenPose generated more false positives than
MTCNN. ViolaJones performed quite poorly relative to the other
detectors, finding much less faces in the random sample.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rllrrr}
  \hline
   Algorithm & Sample\ Type & P & R & F \\ 
  \hline
  MTCNN-Faces & High density & 0.89 & 0.92 & \textbf{0.90} \\ 
  MTCNN-Faces & Random & 0.94 & 0.62 & 0.75 \\ 
  OpenPose-Faces & High density & 0.78 & 0.93 & 0.84 \\ 
  OpenPose-Faces & Random & 0.72 & 0.80 & \textbf{0.76} \\ 
  ViolaJones-Faces & High density & 0.96 & 0.44 & 0.60 \\ 
  ViolaJones-Faces & Random & 0.44 & 0.38 & 0.41 \\ 
  OpenPose-Wrists & High density & 0.64 & 1.00 & 0.78 \\ 
  OpenPose-Wrists & Random & 0.76 & 0.26 & 0.39 \\ 
   \hline
\end{tabular}
\caption{Detector performance on both high density samples (where proportion of faces detected was high) and random samples (where frames were randomly selected). P, R, and F denote precision, recall, and F-score, respectively. Scores in bold are the highest F-scores for each sample type.} 
\end{table}

@@

First, we report developmental shifts in infants' posture and their
orientation relative to their caregiver. Then, we explore how these
changes influence children's visual access to faces and hands across
this developmental time range. Finally, we explore how these changes
impact the accessibility of faces and hands during labeling events.

@@

\section{Changes in Access to Faces and
Hands}\label{changes-in-access-to-faces-and-hands}

@@

\caption[Proportion of faces detected by the MTCNN model (left) and wrists detected by the OpenPose model (right) as a function of child's age]{Proportion of faces detected by the MTCNN model (left) and wrists detected by the OpenPose model (right) as a function of child's age. Larger dots indicate children who had longer play sessions and thus for whom there was more data.}\label{fig:detByAge}
\end{figure}
\end{CodeChunk}

We first examined the proportion of face and wrist detections across
age; a full summary can be seen in Figure \ref{fig:detByAge}. When
looking at face presence, we observed a slight U-shaped function both
when analyzing the output of the MTCNN and OpenPose detectors, such that
12-month-olds appeared to experience slighly fewer faces faces than 8 or
16-month-olds.

@@

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
Intercept & -5.2469 & 0.0584 & -89.88 & 0.0000 \\ 
  Age & 0.0847 & 0.0041 & 20.89 & 0.0000 \\ 
  Prone & 0.2015 & 0.0564 & 3.57 & 0.0004 \\ 
  Sit & 1.4053 & 0.0541 & 25.98 & 0.0000 \\ 
  Stand & 1.4272 & 0.0542 & 26.33 & 0.0000 \\ 
  Close & 1.8239 & 0.0230 & 79.17 & 0.0000 \\ 
  Far & 2.5479 & 0.0239 & 106.42 & 0.0000 \\ 
   \hline
\end{tabular}
\caption{Model coefficients from a generalized linear model predicting the proportion of faces seen by infants.} 
\end{table}

@@

To formalize these observations, we fit a generalized linear model to
the proportion of faces infants saw in each posture and orientation (as
detected by MTCNN), with participant's age, orientation, and posture as
independent variables. A summary of the coefficients of a model with
only main effects (and no interactions) model can be found in Table 3.
When we did include interaction terms between age, posture, and
orientation, age no longer remained a significant predictor (b = -15.84,
SE = 113.98, z = -0.14 , p =0.89). We found the same pattern of results
for the OpenPose face detections, omitted here for brevity. We also
found the same pattern with respect to infant's visual access to hands
(see Table 4); including interaction terms also reduced any main effect
of age on the proportion of hands detected (b = -15.7, SE = 113.98, z =
-0.14 , p =0.89). Thus, these results suggest that infants' access to
social information is heavily influenced by their postural and
locomotive development.

@@

Finally, we analyzed how face and wrists detections changed during
object labeling events as a function of infant's posture and
orientation. Specfically, we analyzed a four-second window around each
labeling event (e.g., ``Look at the {[}zem{]}!''); these labeling events
were hand-annotated and automatically synchronized with the
frame-by-frame face detections. We again found that infants' posture and
orientations impacted the degree to which they saw their caregiver's
face and wrist during a labeling event; infants who were sitting or
standing were more likely to have access to this social information.
